Read Me

### 2023-04-21 code/analysis_code/Race/p1_EDA_race.Rmd
more stuff

```{r message=FALSE, warning=FALSE}
library(here)
library(spdep)
library(GGally)
library(skimr)
library(ggfortify)
library(MASS)
library(ggcorrplot)
library("FactoMineR")
library(tidyverse)
```
https://github.com/ubikuity/List-of-neighboring-states-for-each-US-state
List-of-neighboring-states-for-each-US-state
```{r}
race_df <- readRDS(here("data/processed_data/race_df_cleaned.rds"))
us_nabs <- readRDS(here("data/processed_data/us_nabs.rds"))

@@

```{r}
autoplot(lm(race_df$`Death Rate (AIDSVu)` ~ 1))
autoplot(lm(race_df$`Death Rate (estimated)` ~ 1))
autoplot(lm(race_df$`Ratio Observed/Expected` ~ 1))
```

```{r}
race_df %>% 
  ggplot() +
  geom_histogram(aes(x = `Death Rate (estimated)`, fill = Race), bins = 10)
shapiro.test(race_df$`Death Rate (estimated)`)

race_df %>% 
  ggplot() +
  geom_histogram(aes(x = `Ratio Observed/Expected`, fill = Race), bins = 10)
shapiro.test(race_df$`Ratio Observed/Expected`)
```
Both histograms show a positively skewed distribution. ROE appears to be the slightest bit more normal.

```{r}
race_df %>% 
  ggplot() +
  geom_histogram(aes(x = log(`Death Rate (estimated)` + 1), fill = Race) , bins = 10)
shapiro.test(log(race_df$`Death Rate (estimated)` + 1))

race_df %>% 
  ggplot() +
  geom_histogram(aes(x = log(`Ratio Observed/Expected` + 1), fill = Race) , bins = 10)
shapiro.test(log(race_df$`Ratio Observed/Expected` + 1))
```
The log transformation of Death Rate better fit, but there is still sufficient evidence to reject normality.

```{r}
race_df_DR <- race_df %>% 
  filter(`Death Rate (estimated)` > 0)
b_DR <-  boxcox(lm(race_df_DR$`Death Rate (estimated)` ~ 1))
lambda_DR <- b_DR$x[which.max(b_DR$y)]

race_df_ROE <- race_df %>% 
  filter(`Ratio Observed/Expected` > 0)

b_ROE <-  boxcox(lm(race_df_ROE$`Ratio Observed/Expected` ~ 1))
lambda_ROE <- b_ROE$x[which.max(b_ROE$y)]
```

@@

Box Cox transformation of DR and results of shapiro test.
```{r}
race_df_DR %>%
  mutate(BOX = box_cox(`Death Rate (estimated)`, lambda_DR)) %>%
  ggplot() + geom_histogram(aes(x= BOX, fill = Race), bins = 10) + facet_wrap(. ~ Race)

race_df_DR <- race_df_DR %>%
  mutate(BOX = box_cox(`Death Rate (estimated)`, lambda_DR))

race_df_DR %>%
  pull(BOX) %>% 
  shapiro.test()
##Reject normality

race_df_DR %>%
  filter(Race == "Black") %>% 
  pull(BOX) %>% 
  shapiro.test()
##Fail to reject normality

race_df_DR %>%
  filter(Race == "White") %>% 
  pull(BOX) %>% 
  shapiro.test()
##Reject normality

race_df_DR %>%
  filter(Race == "Hispanic") %>% 
  pull(BOX) %>% 
  shapiro.test()
##Fail to reject normality

race_df_DR %>%
  filter(Race == "Multiple") %>% 
  pull(BOX) %>% 
  shapiro.test()
##Fail to reject normality

par(mfrow=c(1,3))
plot(lm(race_df_DR$BOX ~ 1))
```
Successful transformation of death rate among of Black, Hispanic, and Multiple racial backgrounds persons.

Box Cox transformation of ROE and results of Shapiro test.
```{r}
race_df_ROE %>%
  mutate(BOX = box_cox(`Ratio Observed/Expected`, lambda_ROE)) %>%
  ggplot() + geom_histogram(aes(x= BOX, fill = Race), bins = 10) + facet_wrap(. ~ Race)

race_df_ROE <- race_df_ROE %>%
  mutate(BOX = box_cox(`Ratio Observed/Expected`, lambda_ROE))

race_df_ROE %>%
  pull(BOX) %>% 
  shapiro.test()

race_df_ROE %>%
  filter(Race == "Black") %>% 
  pull(BOX) %>% 
  shapiro.test()

race_df_ROE %>%
  filter(Race == "White") %>% 
  pull(BOX) %>% 
  shapiro.test()

race_df_ROE %>%
  filter(Race == "Hispanic") %>% 
  pull(BOX) %>% 
  shapiro.test()

race_df_ROE %>%
  filter(Race == "Multiple") %>% 
  pull(BOX) %>% 
  shapiro.test()

par(mfrow=c(1,3))
plot(lm(race_df_ROE$BOX ~ 1))
```

Let's save the dataframe for ROE since after the Box Cox transformation for all races was better with ROE than DR.
```{r}
saveRDS(race_df_ROE, here("data/processed_data/race_new.rds"))
```

>>>>

```{r message=FALSE, warning=FALSE}
library(here)
library(spdep)
library(GGally)
library(skimr)
library(ggfortify)
library(MASS)
library(ggcorrplot)
library("FactoMineR")
library(tidyverse)
library(broom)
library(knitr)
library(ggpubr)
```
https://github.com/ubikuity/List-of-neighboring-states-for-each-US-state
List-of-neighboring-states-for-each-US-state
```{r}
race_df <- readRDS(here("data/processed_data/race_df_cleaned.rds"))
us_nabs <- readRDS(here("data/processed_data/us_nabs.rds"))

@@

Let's start by looking at the overall HIV-related death rate by race in the United States.
```{r}
race_df %>% 
  group_by(Race) %>% 
  summarize(tot_deaths = sum(`Death Cases`),
            tot_population = sum(`Population`),
            overall_rate = round(tot_deaths/ tot_population *1e5,1)) %>% 
  ungroup() %>% 
  mutate(percent_by_race = round(tot_deaths/sum(tot_deaths)*100, 1),
         ratio = round(overall_rate/overall_rate[Race =="White"],1)) %>%
  kable(col.names = c("Race/ Ethnicity",
                      "HIV-Related Deaths (n)",
                      "Population (>12 years old)",
                      "Death Rate (per 100,000)",
                      "Percent",
                      "Ratio"))
# using white persons as reference group
```

We can seen that Black persons accounts for the 43.5% of HIV-related deaths in United States, and the rate of HIV-related deaths among this population was 7.2 times high than the rate of HIV-related deaths among White persons.

Next, let's take a look at our response variable
Univariate Analysis

Let's take a look at some box plots of death rate by races and state. 
We'll start by looking at death rate given by the AIDSVu dataset. 
```{r}
race_df %>% 
  ggplot() +
  geom_boxplot(aes(y = `Death Rate (AIDSVu)`, x = Race)) + 
  labs(title = "HIV-Related Death Rate (AIDSVu) by Race/ Ethnicity, United States, 2020", y = "Rate (per 100,000)", x = "Race/ Ethnicity") + theme(legend.position = "none")
```
It seems like something might be strange about the death rates for persons of multiple races. Let's compare to the death rate that we estimated.

```{r}
race_df %>% 
  ggplot() +
  geom_boxplot(aes(y = `Death Rate (estimated)`, x = Race)) + 
  labs(title = "HIV-Related Death Rate (calculated) by Race/ Ethnicity, United States, 2020", y = "Rate (per 100,000)", x = "Race/ Ethnicity") + theme(legend.position = "none")
```
The plots differ drastically for persons of multiple race. It is difficult to tell what might have caused this drastic change, but my inituition tells me there is something strange going on.

Unfortunately, for the sake of this analysis I believe it might be best to remove persons of multiple race/ ethnicities. 
```{r}
race_df <- race_df %>% 
  filter(Race != "Multiple")
```

```{r}
race_df %>% 
  ggplot() +
  geom_boxplot(aes(y = `Death Rate (AIDSVu)`, x = Race)) + 
  labs(title = "HIV-Related Death Rate (AIDSVu) by Race/ Ethnicity, United States, 2020", y = "Rate (per 100,000)", x = "Race/ Ethnicity") + theme(legend.position = "none")

race_df %>% 
  ggplot() +
  geom_boxplot(aes(y = `Death Rate (estimated)`, x = Race)) + 
  labs(title = "HIV-Related Death Rate (calculated) by Race/ Ethnicity, United States, 2020", y = "Rate (per 100,000)", x = "Race/ Ethnicity") +
  theme(legend.position = "none")

ggsave(here("results/dr_boxplot.png"), units="in", width=10, height=7, dpi=300)
```
These plots are quite similar, so I feel good about using the counts and population data in our later analyses.

Let's take a look at the distribution of HIV-related death rate. For the sake of comparison, we will use the estimated HIV death rate as our response variable. 
```{r}
race_df %>% 
  ggplot() +
  geom_histogram(aes(x = `Death Rate (estimated)`, group = Race), bins = 10) +
  facet_grid(. ~ Race) + 
  theme(legend.position = "none") +
  labs(title = "Histogram of Raw Death Rate (estimated)", y = "n", x = "Rate (per 100,000)")
ggsave(here("results/dr_hist.png"), units="in", width=10, height=7, dpi=300)
```
The distribution for Hispanic and White persons is definitely not normally distributed. The distribution for Black persons seems to be roughly normal. Let's subset to test.

```{r}
black_df <- race_df %>% 
  filter(Race == "Black")

hispanic_df <- race_df %>% 
  filter(Race == "Hispanic")

white_df <- race_df %>% 
  filter(Race == "White")
```

```{r}
shapiro.test(black_df$`Death Rate (estimated)`)
```
We fail to reject normality. No transformation of death rate for Black persons is needed.

```{r}
shapiro.test(hispanic_df$`Death Rate (estimated)`)
# reject normality,

#let's try some transformations
shapiro.test(log(hispanic_df$`Death Rate (estimated)` + 1))

hispanic_df  <- hispanic_df %>% 
  mutate(`Death Rate (log transformed)` = log(`Death Rate (estimated)` + 1))
```
Log transformation was able to normalize the death rate distribution for Hispanic persons.

Moving on to the transformation for the death rate of White persons.
```{r}
shapiro.test(log(white_df$`Death Rate (estimated)` + 1))
```
Log transformation was not sufficient. We will have to try a more powerful transformation method.

Let's try a Box Cox transformation. Box Cox requires variables to be greater than 0, so we'll have to subset our dataframe to proceed with this analysis. We will have to add these observations back later. 
```{r}
b_DR <-  boxcox(lm(white_df$`Death Rate (estimated)` ~ 1))

lambda_DR <- b_DR$x[which.max(b_DR$y)] #pulling lambda
```

@@

Box Cox transformation of DR and results of shapiro test.
```{r}
white_df <- white_df %>%
  mutate(`Death Rate (BOX transformed)` = box_cox(`Death Rate (estimated)`, lambda_DR))

white_df %>% 
  pull(`Death Rate (BOX transformed)`) %>% 
  shapiro.test()
```
There is still sufficient efficient to reject normality. Let's see if a more powerful transformation works...

We'll try a Yeo-Johnson transformation.
```{r}
yeo_johnson <- function(x, lambda) {

    eps <- .000001
    not_neg <- which(x >= 0)
    is_neg  <- which(x < 0)

    not_neg_trans <- function(x, lambda) {
        if (abs(lambda) < eps) log(x + 1)
        else ((x + 1) ^ lambda - 1) / lambda
    }

    neg_trans <- function(x, lambda) {
        if (abs(lambda - 2) < eps) - log(-x + 1)
        else - ((-x + 1) ^ (2 - lambda) - 1) / (2 - lambda)
    }

    x[not_neg] <- not_neg_trans(x[not_neg], lambda)

    x[is_neg] <- neg_trans(x[is_neg], lambda)

    return(x)

}

```

```{r}
white_df <- white_df %>%
  mutate(`Death Rate (YJ transformed)` = yeo_johnson(`Death Rate (estimated)`, lambda_DR))

white_df %>% 
  pull(`Death Rate (YJ transformed)`) %>% 
  shapiro.test()
```
Still unable to normalize. We will have to utilize methods that are not sensitive to non-normal distributions for White persons.

```{r}
df <- list(black_df, hispanic_df, white_df)

df <- reduce(df, full_join)

df <- df %>% 
  mutate(outcome_DR = case_when(
    Race == "Black" ~ `Death Rate (estimated)`,
    Race == "White" ~ `Death Rate (estimated)`,
    Race == "Hispanic" ~ `Death Rate (log transformed)`
  ))
```

Let's move on to
Bivariate Analysis
```{r}
ggplot(df, aes(y = outcome_DR, x = `Percent Living in Poverty`)) +
  geom_smooth(method = "lm") + 
  geom_point() + 
  labs(title = "HIV Death Rate vs. Percent Living in Poverty") + 
  stat_cor(method = "pearson") +
  facet_wrap(Race ~ . , scales = "free_y")
ggsave(here("results/bi_pov.png"), units="in", width=10, height=7, dpi=300)
#did not "pass"

ggplot(df, aes(y = outcome_DR, x = `Percent Unemployed`)) +
  geom_smooth(method = "lm") +
  geom_point() + 
  labs(title = "HIV Death Rate vs. Percent Unemployed") + 
  stat_cor(method = "pearson") +
  facet_wrap(Race ~ . , scales = "free_y")
ggsave(here("results/bi_unemployed.png"), units="in", width=10, height=7, dpi=300)
#"passed"

ggplot(df, aes(y = outcome_DR, x = `Gini Coefficient`)) +
  geom_smooth(method = "lm") +
  geom_point() + 
  labs(title = "HIV Death Rate vs. Gini Coefficient") +
  facet_wrap(Race ~ . , scales = "free_y") + 
  stat_cor(method="pearson") 
ggsave(here("results/bi_gini.png"), units="in", width=10, height=7, dpi=300)
#"passed"

ggplot(df, aes(y = outcome_DR, x = `Percent High School Education`)) +
  geom_smooth(method = "lm") + 
  geom_point() + 
  labs(title = "HIV Death Rate vs. Percent High School Education") +
  facet_wrap(Race ~ . , scales = "free_y") +
  stat_cor(method="pearson")
ggsave(here("results/bi_ed.png"), units="in", width=10, height=7, dpi=300)
#"passed"

ggplot(df, aes(y = outcome_DR, x = `Percent Living with Severe Housing Cost Burden`)) +
  geom_smooth(method = "lm") +
  geom_point() + 
  labs(title = "HIV Death Rate  vs. Percent Living with Severe Housing Cost Burden") +
  facet_wrap(Race ~ . , scales = "free_y")+ 
  stat_cor(method="pearson")
#"passed"
ggsave(here("results/bi_housing.png"), units="in", width=10, height=7, dpi=300)

ggplot(df, aes(y = outcome_DR, x = `Broadband of any type`)) +
  geom_smooth(method = "lm") +
  geom_point() + 
  labs(title = "HIV Death Rate vs. Broadband of any type") +
  facet_wrap(Race ~ . , scales = "free_y") + 
  stat_cor(method="pearson")
# did not "pass"
```

```{r}
ggplot(df, aes(y = outcome_DR, x = `Percent Uninsured Under 65 Years Old`)) +
  geom_smooth(method = "lm") +
  geom_point() + 
  labs(title = "HIV Death Rate vs. Percent Uninsured Under 65 Years Old") +
  facet_wrap(Race ~ . , scales = "free_y") + 
  stat_cor(method="pearson")
# did not "pass"

ggplot(df, aes(y = outcome_DR, x = `Percent Living with Food Insecurity`)) +
  geom_smooth(method = "lm") +
  geom_point() + 
  labs(title = "HIV Death Rate vs. Percent Living with Food Insecurity") +
  facet_wrap(Race ~ . , scales = "free_y") + 
  stat_cor(method="pearson")
# did not "pass"

ggplot(df, aes(y = outcome_DR, x = `Broadband of any type`)) +
  geom_smooth(method = "lm") +
  geom_point() + 
  labs(title = "HIV Death Rate vs. Broadband of any type") +
  facet_wrap(Race ~ . , scales = "free_y") + 
  stat_cor(method="pearson")
# did not "pass"

ggplot(df, aes(y = outcome_DR, x = `Median Household Income`)) +
  geom_smooth(method = "lm") +
  geom_point() + 
  labs(title = "HIV Death Rate vs. Median Household Income") +
  facet_wrap(Race ~ . , scales = "free_y") + 
  stat_cor(method="pearson")
# did not "pass"

ggplot(df, aes(y = outcome_DR, x = `Percentage Diagnosed Late`)) +
  geom_smooth(method = "lm") +
  geom_point() + 
  labs(title = "HIV Death Rate vs. Percentage Diagnosed Late") +
  facet_wrap(Race ~ . , scales = "free_y") + 
  stat_cor(method="pearson")

ggplot(df, aes(y = outcome_DR, x = `Percentage Virally Suppressed`)) +
  geom_smooth(method = "lm") +
  geom_point() + 
  labs(title = "HIV Death Rate vs. Percentage Virally Suppressed") +
  facet_wrap(Race ~ . , scales = "free_y") + 
  stat_cor(method="pearson")
```
Now that we have an idea of the relationship between our outcome and potential predictors, next I need to account for the fact that these variables a likely have some degree of multicollinearity. In the next step we will be using PCA to scale our predictors.

```{r}
saveRDS(df, here("data/processed_data/race_new.rds"))
```

### 2023-04-21 code/analysis_code/Race/p2_PCA_race.Rmd
more stuff

```{r message=FALSE, warning=FALSE}
library(ggcorrplot)
library(FactoMineR)
library(factoextra)
library(spatialreg)
library(MASS)
library(here)
library(tidyverse)

@@

```{r}
pca_function <- function(race) {   
  # filter race_df by the selected race/ethnicity
  df <- df[df$Race == race,]

num_race <- df[,c(9:10, 16:25)]

#normalization
norm_race <- scale(num_race)
print(head(norm_race))

#setting up correlation matrix
corr_matrix_race <- cor(norm_race)
print(ggcorrplot(corr_matrix_race))

#applying pca
pca_race <- princomp(corr_matrix_race)
print(summary(pca_race))
#pca loadings
print(pca_race$loadings[, 1:2])

#Scree plot
print(fviz_eig(pca_race, addlabels = TRUE))
# Biplot of the attributes
print(fviz_pca_var(pca_race, col.var = "darkblue"))

#Contribution of each variable 
print(fviz_cos2(pca_race, choice = "var", axes = 1:2))

# Biplot of the attributes
print(fviz_pca_var(pca_race, col.var = "cos2",
             gradient.cols = c("darkblue", "orange", "green"),
             repel = TRUE))
# Variables in green are a good representation of the variable on that component.
}
```

```{r}
pca_function("Black")
```

```{r}
pca_function("Hispanic")
```

```{r}
pca_function("Multiple")
```

```{r}
pca_function("White")
```

>>>>

```{r message=FALSE, warning=FALSE}
library(ggcorrplot)
library(FactoMineR)
library(factoextra)
library(here)
library(knitr)
library(tidyverse)
library(broom)

@@

PCA
```{r}
df <- df %>%
    select_if(~ !any(is.na(.)))

  #separating the y variables
df_response <- df %>%
    select(contains(c("State", "Race", "Death", "Population", "BOX", "YJ", "Ratio", "outcome_DR")))

  # separating x variables and scaling
df_predictor <- df %>% 
    select(contains(c("State", "Race", "Percent", "Gini", "Median", "Broadband"))) %>% 
    #select(!contains(c("Virally", "Received", "Late"))) %>% 
    mutate(across(where(is.numeric), scale))

#Setting up correlation matrix
corr_matrix <- df_predictor %>% 
    select(where(is.numeric)) %>% 
    cor()

  #setting up correlation matrix 
print(ggcorrplot(corr_matrix))
ggsave(here("results/corr_mat.png"), units="in", width=10, height=7, dpi=300)
```

```{r}
#applying pca
pca <- princomp(corr_matrix)
print(summary(pca))

#pca loadings of principal components 1 and 2
print(pca$loadings[, 1:2])

#Scree plot
print(fviz_eig(pca, addlabels = TRUE))
ggsave(here("results/scree_plot.png"), units="in", width=10, height=7, dpi=300)
```
Since we originally had 11 variables, we should only look at components that explain more than 9.1% of variation. We can see from the scree plot that principal component 1 explained 63.8% of variation, and components 1 and 2 explained 84.7% of the variation, and components 1 - 3 explained 94.8%. Component 3 just  barely passed the threshold with 10.1%. All other principal components explains less that 6.2% of variation, so they will excluded.

```{r}
#Contribution of each variable 
print(fviz_cos2(pca, choice = "var", axes = 1:2))

# Biplot of the attributes
print(fviz_pca_var(pca, col.var = "cos2",
                   gradient.cols = c("darkblue", "orange", "green"),
                   repel = TRUE, title = "PCA"
                   ))
# Variables in green are a good representation of the variable on that component.

#Creating object to store scaled dataframes
df_scaled <- full_join(df_response, df_predictor, by = join_by("State", "State Abbreviation", "States with HIV-Specific Criminal Laws/Statutes", "States That Have Prosecuted PLWH", "Race"))
```

```{r}
saveRDS(df_scaled, here("data/processed_data/df_scaled.rds"))
```

### 2023-04-21 code/analysis_code/Race/p3_model_race.Rmd
more stuff

```{r message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(spdep)
```

```{r}
df <- readRDS(here("data/processed_data/race_new.rds"))
us_nabs <- readRDS(here("data/processed_data/us_nabs.rds"))

@@

df <- df %>% 
  filter(!grepl("Alaska", State),
         !grepl("Hawaii", State)) %>% 
  mutate(log_rate = log(`Death Rate (estimated)`+ 1))
```

Setting up functions for models
```{r}
  #poisson
pois_function <- function(race) {
  pois_race <- glm(`Death Cases` ~ `Median Household Income` +
                       `Percent High School Education` +
                       offset(log(`Population`)),
                     data = df[df$Race == race,],
                     family = poisson)
  print(summary(pois_race))
}

  #Negative binomial
nb_function  <- function(race) {
  neg_bin_race <- glm.nb(`Death Cases` ~ `Median Household Income` +
                              `Percent High School Education` + 
                              offset(log(`Population`)),
                            data = df[df$Race == race,])
  print(summary(neg_bin_race))
}

  #Fitting CAR model (requires neighbors)
car_function <- function(race) {   
  # filter race_df by the selected race/ethnicity
  df <- df[df$Race == race,]

@@

  car_race <- spautolm(log_rate ~ `Percent Living in Poverty` + 
                             `Percent High School Education`,
                           data = df,
                           listw = nablist_race_df,
                           family = "CAR")
  print(summary(car_race))
}
```

Fitting Poisson models
```{r}
pois_function("Black")
```

```{r}
pois_function("Hispanic")
```

```{r}
pois_function("Multiple")
```

```{r}
pois_function("White")
```

Fitting Negative binomial models
```{r}
nb_function("Black")
```

```{r}
nb_function("Hispanic")
```

```{r}
nb_function("Multiple")
```

```{r}
nb_function("White")
```
The NB model accounting for overdispersion fit better than the poisson model.

Fitting CAR models
```{r}
car_function("Black")
```

```{r}
car_function("Hispanic")
```

```{r}
car_function("Multiple")
```

```{r}
car_function("White")
```
The model accounting for spatial variation provide the best fit.

Creating function that subsets by specified race, and compute/ plots Moran's I and lisa.
```{r}
spacy_function <- function(race) {
  # filter race_df by the selected race/ethnicity
  df <- df[df$Race == race,]

 # plot Moran's I
  par(mfrow=c(1,2))
  print(moran.test(df$BOX, nablist_race_df, zero.policy = TRUE, randomisation = FALSE))
  moran.plot(df$BOX, nablist_race_df, zero.policy = TRUE)

  # calculate and print LISA statistics
  lisa_race <- localmoran(df$`Ratio Observed/Expected`, nablist_race_df)
  summary(lisa_race[,1])
}
```

```{r}
spacy_function("Black")
```

```{r}
spacy_function("Hispanic")
```

```{r}
spacy_function("Multiple")
```

```{r}
spacy_function("White")
```

```{r}
save.image(file="../../data/processed_data/race_env.rds")
```

>>>>

```{r message=FALSE, warning=FALSE}
library(here)
library(MASS)
library(spatialreg)
library(spdep)
library(tidyverse)
```

```{r}
df <- readRDS(here("data/processed_data/df_scaled.rds"))
us_nabs <- readRDS(here("data/processed_data/us_nabs.rds"))

@@

df <- df %>% 
  filter(!grepl("Alaska", State),
         !grepl("Hawaii", State))
```

subset by race/ethnicity
```{r}
black_df <- df %>% 
  filter(Race == "Black")

hispanic_df <- df %>% 
  filter(Race == "Hispanic")

white_df <- df %>% 
  filter(Race == "White")
```

Setting up functions for models

```{r}
 #multiple linear regression
mult_function <- function(df) {
  mult_race <- glm(outcome_DR ~ 
                        `Gini Coefficient` + # estimate income inequality
                        `Percent Unemployed` +
                        `Percent Living in Poverty` +
                        `Percent High School Education` +  #estimate education access
                 #       `Broadband of any type` + # estimate built 
                   #     `Percent Uninsured Under 65 Years Old`+ #estimate access to healthcare
                        `Percent Living with Severe Housing Cost Burden`, #economic stability,
                     data = df,
                     family = "gaussian")
  print(summary(mult_race))
    ##pull fitted values
df$mult_fit <- mult_race$fitted.values
return(df)
}

## function to fit GLM and pull AIC
mult_aic <- function(df) {
  mult_race <- glm(outcome_DR ~ 
                        `Gini Coefficient` + # estimate income inequality
                        `Percent Unemployed` +
                     `Percent Living in Poverty` +
                        `Percent High School Education` +  #estimate education access
                    #     `Broadband of any type` + # estimate built 
                     #   `Percent Uninsured Under 65 Years Old`+ #estimate access to healthcare
                        `Percent Living with Severe Housing Cost Burden`, #economic stability,
                     data = df,
                     family = "gaussian")
aic <- AIC(mult_race)

return(aic)
}
```

```{r}
  #poisson
pois_function <- function(df) {
  pois_race <- glm(`Death Cases`~ 
                        `Gini Coefficient` + # estimate income inequality
                        `Percent Unemployed` +
                     `Percent Living in Poverty` +
                        `Percent High School Education` +  #estimate education access
                      #   `Broadband of any type` + # estimate built 
                     #   `Percent Uninsured Under 65 Years Old` + #estimate access to healthcare
                        `Percent Living with Severe Housing Cost Burden` + #economic stability
                    offset(log(`Population`)),
                     data = df,
                     family = "poisson")
  print(summary(pois_race))

      ##pull fitted values
df$pois_fit <- pois_race$fitted.values
return(df)
}

## function to fit GLM and pull AIC
pois_aic <- function(df) {
  pois_race <- glm(`Death Cases`~ 
                        `Gini Coefficient` + # estimate income inequality
                        `Percent Unemployed` +
                     `Percent Living in Poverty` +
                        `Percent High School Education` +  #estimate education access
                     #    `Broadband of any type` + # estimate built 
                    #    `Percent Uninsured Under 65 Years Old` + #estimate access to healthcare
                        `Percent Living with Severe Housing Cost Burden` + #economic stability
                    offset(log(`Population`)),
                     data = df,
                     family = "poisson")
aic <- AIC(pois_race)
return(aic)
}
```

```{r}
  #Negative binomial
nb_function  <- function(df) {
  neg_bin_race <- glm.nb(`Death Cases` ~ 
                        `Gini Coefficient` + # estimate income inequality
                        `Percent Unemployed` +
                          `Percent Living in Poverty` +
                        `Percent High School Education` +  #estimate education access
                     #   `Broadband of any type` + # estimate built 
                      #  `Percent Uninsured Under 65 Years Old` + #estimate access to healthcare
                        `Percent Living with Severe Housing Cost Burden` + #economic stability
                     offset(log(`Population`)),
                            data = df)
  print(summary(neg_bin_race))

      ##pull fitted values
df$nb_fit <- neg_bin_race$fitted.values

return(df)

}

## function to fit GLM and pull AIC
nb_aic <- function(df) {
  neg_bin_race <- glm.nb(`Death Cases` ~ 
                        `Gini Coefficient` + # estimate income inequality
                        `Percent Unemployed` +
                          `Percent Living in Poverty` +
                        `Percent High School Education` +  #estimate education access
                      #   `Broadband of any type` + # estimate built 
                   #    `Percent Uninsured Under 65 Years Old` + #estimate access to healthcare
                        `Percent Living with Severe Housing Cost Burden` + #economic stability
                     offset(log(`Population`)),
                            data = df)
aic <- AIC(neg_bin_race)

return(aic)

}  
```

```{r}
#Fitting CAR model (requires neighbors)
car_function <- function(df) {   
  h_race_df <- matrix(0, nrow(df), nrow(df))
  h_race_df[cbind(match(us_nabs$StateCode, df$`State Abbreviation`),
                  match(us_nabs$NeighborStateCode, df$`State Abbreviation`))] <- 1

  car_race <- spautolm(outcome_DR ~                      
                        `Gini Coefficient` + # estimate income inequality
                        `Percent Unemployed` +
                         `Percent Living in Poverty` +
                        `Percent High School Education` +  #estimate education access
                       #  `Broadband of any type` + # estimate built 
                     #   `Percent Uninsured Under 65 Years Old`+ #estimate access to healthcare
                        `Percent Living with Severe Housing Cost Burden`,#economic stability,
                           data = df,
                           listw = nablist_race_df,
                           family = "CAR")
  print(summary(car_race))

  ##pull fitted values
df$car_fit <- car_race$fit$fitted.values

return(df)

}

car_aic <- function(df) {
  h_race_df <- matrix(0, nrow(df), nrow(df))
  h_race_df[cbind(match(us_nabs$StateCode, df$`State Abbreviation`),
                  match(us_nabs$NeighborStateCode, df$`State Abbreviation`))] <- 1

  car_race <- spautolm(outcome_DR ~                      
                        `Gini Coefficient` + # estimate income inequality
                        `Percent Unemployed` +
                         `Percent Living in Poverty` +
                        `Percent High School Education` +  #estimate education access
                     #    `Broadband of any type` + # estimate built 
                    #    `Percent Uninsured Under 65 Years Old`+ #estimate access to healthcare
                        `Percent Living with Severe Housing Cost Burden`,#economic stability,
                           data = df,
                           listw = nablist_race_df,
                           family = "CAR")
aic <- AIC(car_race)
return(aic)

} 
```

Making a function to pull AIC and organize into new dataframe
```{r}
aic_df <- function(x, mod_type, race) {
  x <- tibble(x)
  x <- x %>% 
    mutate(Race = race, 
           Model = mod_type,
           AIC = round(x,2)) %>% 
    select(Race, Model, AIC)
}
```

Fitting multiple linear regression models
```{r}
black_df <- mult_function(black_df)

black_mult <- mult_aic(black_df)
black_mult <- aic_df(black_mult, "GLM", "Black")
```

```{r}
hispanic_df <- mult_function(hispanic_df)

hispanic_mult <- mult_aic(hispanic_df)
hispanic_mult <- aic_df(hispanic_mult, "GLM", "Hispanic")
```

```{r}
white_df <- mult_function(white_df)
## SINCE THE OUTCOME IN NOT NORMALLY DISTRIBUTED, WE CANNOT USE GLM 
white_df <- white_df %>% mutate(mult_fit = NA)

white_mult <- mult_aic(white_df)
white_mult <- aic_df(white_mult, "GLM", "White")

white_mult <- white_mult %>% mutate(AIC = NA)
```

```{r}
mult_mod <- list(black_mult, hispanic_mult, white_mult)

mult_mod <- reduce(mult_mod, full_join)
```

Fitting Poisson models
```{r}
black_df <- pois_function(black_df)

black_pois <- pois_aic(black_df)
black_pois <- aic_df(black_pois, "Poisson", "Black")
```

```{r}
hispanic_df <- pois_function(hispanic_df)

hispanic_pois <- pois_aic(hispanic_df)
hispanic_pois <- aic_df(hispanic_pois, "Poisson", "Hispanic")
```

```{r}
white_df <- pois_function(white_df)

white_pois <- pois_aic(white_df)
white_pois <- aic_df(white_pois, "Poisson", "White")
```

Storing Poisson AIC
```{r}
pois_mod <- list(black_pois, hispanic_pois, white_pois)

pois_mod <- reduce(pois_mod, full_join)
```

Fitting Negative binomial models
```{r}
black_df <- nb_function(black_df)

black_nb <- nb_aic(black_df)
black_nb <- aic_df(black_nb, "Negative Binomial", "Black")
```

```{r}
hispanic_df <- nb_function(hispanic_df)

hispanic_nb <- nb_aic(hispanic_df)
hispanic_nb <- aic_df(hispanic_nb, "Negative Binomial", "Hispanic")
```

```{r}
white_df <- nb_function(white_df)

white_nb <- nb_aic(white_df)
white_nb <- aic_df(white_nb, "Negative Binomial", "White")
```

```{r}
nb_mod <- list(black_nb, hispanic_nb, white_nb)

nb_mod <- reduce(nb_mod, full_join)
```
The NB model accounting for overdispersion fits better than the poisson model.

Fitting CAR models
```{r}
black_df <- car_function(black_df)

black_car <- car_aic(black_df)
black_car <- aic_df(black_car, "CAR", "Black")
```

```{r}
hispanic_df <- car_function(hispanic_df)

hispanic_car <- car_aic(hispanic_df)
hispanic_car <- aic_df(hispanic_car, "CAR", "Hispanic")
```

```{r}
white_df <- car_function(white_df)
## SINCE THE OUTCOME IN NOT NORMALLY DISTRIBUTED, WE CANNOT USE CAR
white_df <- white_df %>% mutate(car_fit = NA)

white_car <- car_aic(white_df)
white_car <- aic_df(white_car, "CAR", "White")
white_car <- white_car %>% mutate(AIC = NA)
```

```{r}
car_mod <- list(black_car, hispanic_car, white_car)

car_mod <- reduce(car_mod, full_join)
```

```{r}
mod_aic <- list(mult_mod, pois_mod, nb_mod, car_mod)

mod_aic <- reduce(mod_aic, full_join)
```

Join the data frames with all the fits.
```{r}
df <- list(black_df, hispanic_df, white_df)

df <- reduce(df, full_join)

saveRDS(df, here("data/processed_data/df_fit.rds"))
```
save the AIC dataframe
```{r}
saveRDS(mod_aic, here("data/processed_data/aic.rds"))
```
The model accounting for spatial variation provide the best fit.

Creating function that subsets by specified race, and compute/ plots Moran's I and lisa.
```{r}
spacy_function <- function(df) {

@@

 # plot Moran's I
  print(moran.test(df$outcome_DR, nablist_race_df, zero.policy = TRUE, randomisation = FALSE))
  moran.plot(df$outcome_DR, nablist_race_df, zero.policy = TRUE)

  # calculate and print LISA statistics
  lisa_race <- localmoran(df$outcome_DR, nablist_race_df)
  summary(lisa_race[,1])
}
```

```{r}
spacy_function(black_df)
```

```{r}
png(here("results/moran_black.png"))
#sink(here("results/spacy_summary_black.txt"))
spacy_function(black_df)
sink()
dev.off()
```

```{r}
spacy_function(hispanic_df)
```

```{r}
png(here("results/moran_hispanic.png"))
#sink(here("results/spacy_summary_hispanic.txt"))
spacy_function(hispanic_df)
sink()
dev.off()
```

```{r}
spacy_function(white_df)
```

```{r}
png(here("results/moran_white.png"))
#sink(here("results/spacy_summary_white.txt"))
spacy_function(white_df)
sink()
dev.off()
```

```{r}
save.image(here("data/processed_data/race_env.rds"))
```
